### 2019-10-16 ch3_dynbenchmark/content.tex
update thesis

A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table 1). Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories. While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}. These methods therefore mainly focus on correctly ordering the cells along the fixed topology. More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_1.pdf}
	\caption{
		\textbf{Overview of several key aspects of the evaluation.}
		\textbf{a}, A schematic overview of our evaluation pipeline. \textbf{b}, To make the trajectories comparable to each other, a common trajectory model was used to represent reference trajectories from the real and synthetic datasets, as well as any predictions of TI methods. \textbf{c}, Trajectories are automatically classified into one of seven trajectory types, with increasing complexity. \textbf{d}, We defined four metrics, each assessing the quality of a different aspect of the trajectory. The HIM score assesses the similarity between the two topologies, taking into account differences in edge lengths and degree distributions. The F1branches assesses the similarity of the assignment of cells onto branches. The cordist quantifies the similarity in cellular positions between two trajectories, by calculating the correlation between pairwise geodesic distances. Finally, wcorfeatures quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory.
	}
	\label{fig:figure_1}
\end{figure}

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/supfigure_1.pdf}
	\caption{
		\textbf{A common interface for TI methods.}
		\textbf{a} The input and output of each TI method is standardized. As input, each TI method receives either raw or normalized counts, several parameters, and a selection of prior information. After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. \textbf{b} Illustrations of the specific transformations performed by each of the wrapper functions.
	}
	\label{fig:supfigure_1}
\end{figure}

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_2.pdf}
	\caption{
		\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
		\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies. The methods are grouped vertically based on the most complex trajectory type they can infer. \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation. Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig. 2.
	}
	\label{fig:figure_2}
\end{figure}

@@

We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript. Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b). We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig. 2), after which we conclude with guidelines for method users and future perspectives for method developers.

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_3.pdf}
	\caption{	
		\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
		\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}. \textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. \textbf{c}, Predicted execution times for varying numbers of cells and features (no. of cells $\times$ no. of features). Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features. k, thousands; m, millions; cor, correlation. \textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method. \textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category. Off-the-shelf methods were directly implemented in R and thus do not have a usability score. 
	}
	\label{fig:figure_3}
\end{figure}

\subsection{Accuracy}

We defined several metrics to compare a prediction to a reference trajectory (Supplementary Note 1). Based on an analysis of their robustness and conformity to a set of rules (Supplementary Note 1), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table 2). Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}. All other real datasets were classified as ‘silver standard’. For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter_genenetweaversilicobenchmark_2011}. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/supfigure_3.pdf}
	\caption{
		\textbf{Accuracy of trajectory inference methods.}
		\textbf{a} Overall score for all methods across 339 datasets, colored by the source of the datasets. Black line indicates the mean. \textbf{b} Similarity between the overall scores of all dataset sources, compared to real datasets with a gold standard, across all methods (n = 46, after filtering out methods that errored too frequently). Shown in the top left is the Pearson correlation. \textbf{c} Bias in the overall score towards trajectory types for all methods across 339 datasets. Black line indicates the mean. \textbf{d} Distributions of the difference in size between predicted and reference topologies. A positive difference means that the topology predicted by the method is more complex than the one in the reference.
	}
	\label{fig:supfigure_3}
\end{figure}

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_4.pdf}
	\caption{
		\textbf{Complementarity between different trajectory inference methods.}
		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95$\%$ of the best model) when applied to all datasets. \textbf{b}, The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types. For this figure, we did not include any methods requiring a cell grouping or a time course as prior information.
	}
	\label{fig:figure_4}
\end{figure}

@@

While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million \cite{svensson_exponentialscalingsinglecell_2018}. Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features \cite{cao_jointprofilingchromatin_2018}. To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets. We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a). As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Supplementary Fig. 5), and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig. 2a).

\begin{figure}[htb!]
	\centering\includegraphics[width=0.9\linewidth]{fig/supfigure_4.pdf}
	\caption{
		\textbf{Scalability of trajectory inference methods.}
		\textbf{a} Three examples of average observed running times across five datasets (left) and the predicted running time (right). \textbf{b} Overview of the scalability results of all methods, ordered by their average predicted running time from (a). We predicted execution times and memory usage for each method with increasing number of features or cells, and used these values to classify each method into sublinear, linear, quadratic and superquadratic based on the shape of the curve.
	}
	\label{fig:supfigure_4}
\end{figure}

@@

We also assessed the memory requirements of each method (Supplementary Fig. 2c). Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features. Notably, the memory requirements were very high for several methods on datasets with high numbers of cells (RaceID/StemID, pCreode and MATCHER) or features (Monocle DDRTree, SLICE and MFA).

Altogether, the scalability analysis indicated that the dimensions of the data are an important factor in the choice of method, and that method development should pay more attention to maintaining reasonable running times and memory usage.
§
\subsection{Stability}

@@

We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Supplementary Fig. 6). While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Supplementary Fig. 6 right) and we believe that these should receive extra attention from developers. Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_bestpracticesscientific_2014,artaza_top10metrics_2016}. Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration for future methods. We observed no clear relation between usability and method accuracy or usability (Figure \ref{fig:figure_2}b).

\begin{table}
	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.} \label{tab:scoresheet}

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_5.pdf}
	\caption{
		\textbf{Practical guidelines for method users.}
		As the performance of a method mostly depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user’s existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method’s performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to their performance on a particular (set of) trajectory type. Further to the right are shown the accuracy (+: scaled performance $\geq$ 0.9, $\pm$: >0.6), usability scores (+:$\geq$0.9, $\pm$ $\geq$0.6), estimated running times and required prior information. k, thousands; m, millions.
	}
	\label{fig:figure_5}
\end{figure}

@@

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_6.pdf}
	\caption{
		\textbf{Demonstration of how a common framework for TI methods facilitates broad applicability using some example datasets.}
		Trajectories inferred by each method were projected to a common dimensionality reduction using multidimensional scaling. For each dataset, we also calculated a ‘consensus’ prediction, by calculating the cordist between each pair of models and picking the model with the highest score on average. \textbf{a}, The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC. \textbf{b}, The top methods applied on a dataset containing a bifurcating trajectory of reprogrammed fibroblasts. \textbf{c}, A synthetic dataset generated by dyntoy, containing four disconnected trajectories. \textbf{d}, A synthetic dataset generated by dyngen, containing a cyclic trajectory.
	}
	\label{fig:figure_6}
\end{figure}

@@

We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}. We welcome any contributions by creating an issue at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.

@@

We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table 2). The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalized and filtered using a common pipeline, as discussed later. Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation. In the end, we included 110 datasets for this evaluation.

@@

The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{norel_selfassessmenttrapcan_2011}. Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer. We therefore defined several possible metrics for comparing trajectories, each investigating different layers. These are all discussed in Supplementary Note 1 along with examples and robustness analyses when appropriate.

Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (Supplementary Note 1). Based on this analysis, we chose four metrics for the evaluation, each assessing a different aspect of the trajectory: (1) the HIM measures the topological similarity; (2) the F1branches compares the branch assignment; (3) the cordist assesses the similarity in pairwise cell–cell distances and thus the cellular positions; and (4) the wcorfeatures looks at whether similar important features (genes) are found in both the reference dataset and the prediction.

@@

The HIM metric \cite{jurman_himglocalmetric_2015} uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length). It is a linear combination of the normalized Hamming distance, which gives an indication of the differences in edge lengths, and the normalized Ipsen–Mikhailov distance, which assesses the similarity in degree distributions. The latter has a parameter $\gamma$, which was fixed at 0.1 to make the scores comparable between datasets. We illustrate the metric and discuss alternatives in Supplementary Note 1.

@@

To compare branch assignment, we used an F1 score, also used used for comparing biclustering methods \cite{saelens_comprehensiveevaluationmodule_2018}. To calculate this metric, we first calculated the similarity of all pairs of branches between the two trajectories using the Jaccard similarity. Next, we defined the ‘Recovery' (respectively ‘Relevance') as the average maximal similarity of all branches in the reference dataset (respectively prediction). The F1branches was then defined as the harmonic mean between Recovery and Relevance. We illustrate this metric further in Supplementary Note 1.

@@

To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics. This aggregation strategy is explained in more detail in Supplementary Note 1.

>>>>

A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}{1}). Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories. While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}. These methods therefore mainly focus on correctly ordering the cells along the fixed topology. More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.

@@

\begin{figure}[tbh!]
	\centering\includegraphics[width=\linewidth]{fig/figure_1.pdf}
	\caption{
		\textbf{Overview of several key aspects of the evaluation.}
		\textbf{a}, A schematic overview of our evaluation pipeline. \textbf{b}, To make the trajectories comparable to each other, a common trajectory model was used to represent reference trajectories from the real and synthetic datasets, as well as any predictions of TI methods. \textbf{c}, Trajectories are automatically classified into one of seven trajectory types, with increasing complexity. \textbf{d}, We defined four metrics, each assessing the quality of a different aspect of the trajectory. The HIM score assesses the similarity between the two topologies, taking into account differences in edge lengths and degree distributions. The F1branches assesses the similarity of the assignment of cells onto branches. The cordist quantifies the similarity in cellular positions between two trajectories, by calculating the correlation between pairwise geodesic distances. Finally, wcorfeatures quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory.
	}
	\label{fig:figure_1}
\end{figure}

@@

\begin{figure}[tbh]
	\centering\includegraphics[width=\linewidth]{fig/supfigure_1.pdf}
	\caption{
		\textbf{A common interface for TI methods.}
		\textbf{a} The input and output of each TI method is standardized. As input, each TI method receives either raw or normalized counts, several parameters, and a selection of prior information. After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. \textbf{b} Illustrations of the specific transformations performed by each of the wrapper functions.
	}
	\label{fig:supfigure_1}
\end{figure}

\afterpage{%
	\clearpage% flush all other floats
	\ifodd\value{page}
%	\else% uncomment this else to get odd/even instead of even/odd
	\expandafter\afterpage% put it on the next page if this one is odd
	\fi
	{%
		\begin{figure}[tbh!]
			\centering\includegraphics[height=.75\textheight]{fig/figure_2.pdf}
			\caption{
				\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
				\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies. The methods are grouped vertically based on the most complex trajectory type they can infer. \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation. Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}.
			}
			\label{fig:figure_2}
		\end{figure}
		\clearpage
		\begin{figure}[tbh!]
			\centering\includegraphics[height=.75\textheight]{fig/figure_3.pdf}
			\caption{	
				\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
				\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}. \textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. \textbf{c}, Predicted execution times for varying numbers of cells and features (no. of cells $\times$ no. of features). Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features. k, thousands; m, millions; cor, correlation. \textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method. \textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category. Off-the-shelf methods were directly implemented in R and thus do not have a usability score. 
			}
			\label{fig:figure_3}
		\end{figure}
		\clearpage
	}%
}

@@

We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript. Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b). We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}), after which we conclude with guidelines for method users and future perspectives for method developers.

\newpage
\subsection{Accuracy}

We defined several metrics to compare a prediction to a reference trajectory (Section~\ref{sec:dynb_supn1}). Based on an analysis of their robustness and conformity to a set of rules (Section~\ref{sec:dynb_supn1}), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}{2}). Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}. All other real datasets were classified as ‘silver standard’. For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter_genenetweaversilicobenchmark_2011}. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.

@@

@@

\begin{figure}[tbh!]
	\centering\includegraphics[width=\linewidth]{fig/supfigure_3.pdf}
	\caption{
		\textbf{Accuracy of trajectory inference methods.}
		\textbf{a} Overall score for all methods across 339 datasets, colored by the source of the datasets. Black line indicates the mean. \textbf{b} Similarity between the overall scores of all dataset sources, compared to real datasets with a gold standard, across all methods (n = 46, after filtering out methods that errored too frequently). Shown in the top left is the Pearson correlation. \textbf{c} Bias in the overall score towards trajectory types for all methods across 339 datasets. Black line indicates the mean. \textbf{d} Distributions of the difference in size between predicted and reference topologies. A positive difference means that the topology predicted by the method is more complex than the one in the reference.
	}
	\label{fig:supfigure_3}
\end{figure}

\begin{figure}[tbh!]
	\centering\includegraphics[width=\linewidth]{fig/figure_4.pdf}
	\caption{
		\textbf{Complementarity between different trajectory inference methods.}
		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95$\%$ of the best model) when applied to all datasets. \textbf{b}, The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types. For this figure, we did not include any methods requiring a cell grouping or a time course as prior information.
	}
	\label{fig:figure_4}
\end{figure}

@@

While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million \cite{svensson_exponentialscalingsinglecell_2018}. Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features \cite{cao_jointprofilingchromatin_2018}. To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets. We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a). As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Figure \ref{fig:supfigure_5}, and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}a).

\begin{figure}[tbh!]
	\centering\includegraphics[width=0.9\linewidth]{fig/supfigure_4.pdf}
	\caption{
		\textbf{Scalability of trajectory inference methods.}
		\textbf{a} Three examples of average observed running times across five datasets (left) and the predicted running time (right). \textbf{b} Overview of the scalability results of all methods, ordered by their average predicted running time from (a). We predicted execution times and memory usage for each method with increasing number of features or cells, and used these values to classify each method into sublinear, linear, quadratic and superquadratic based on the shape of the curve.
	}
	\label{fig:supfigure_4}
\end{figure}

\begin{figure}[tbh!]
	\centering\includegraphics[width=0.6\linewidth]{fig/supfigure_5.pdf}
	\caption{
		\textbf{Agreement between actual values and predictions for execution times and memory usage.}
		We created a predictive model of the running time and memory usage based on a set of scaling datasets (left), and validated this model based on the similarity of the predictions and actual values on all benchmark datasets (right). Shown are the values for each method and dataset (n = 65618 for training, n = 11939 for test). Top left indicates the Pearson correlation coefficient.
	}
	\label{fig:supfigure_5}
\end{figure}

@@

We also assessed the memory requirements of each method (Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}c). Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features. Notably, the memory requirements were very high for several methods on datasets with high numbers of cells (RaceID/StemID, pCreode and MATCHER) or features (Monocle DDRTree, SLICE and MFA).

Altogether, the scalability analysis indicated that the dimensions of the data are an important factor in the choice of method, and that method development should pay more attention to maintaining reasonable running times and memory usage.

@@

We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Figure~\ref{fig:supfigure_6}). While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Figure~\ref{fig:supfigure_6} right) and we believe that these should receive extra attention from developers. Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_bestpracticesscientific_2014,artaza_top10metrics_2016}. Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration for future methods. We observed no clear relation between usability and method accuracy or usability (Figure \ref{fig:figure_2}b).

\begin{figure}[tbh!]
	\centering\includegraphics[width=0.9\linewidth]{fig/supfigure_6.pdf}
	\caption{
		\textbf{Usability of trajectory inference methods.}
		Shown is the score given for each method on every item from the usability score sheet(Table~\ref{tab:scoresheet}). Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score. Within each  category,  each  aspect  also  received  a  weight  depending  on  how  often  it  was mentioned  in  a set  of  papers  discussing  good practices in tool development and evaluation. This is represented in the plot as the height on the y-axis. Top: Average usability score for each method. Right: The average score of each quality control item. Shown into more detail are those items which had an average score lower than 0.5.
	}
	\label{fig:supfigure_6}
\end{figure}

\begin{table}[tbh!]
	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.} \label{tab:scoresheet}

@@

\begin{figure}[tbh!]
	\centering\includegraphics[width=\linewidth]{fig/figure_5.pdf}
	\caption{
		\textbf{Practical guidelines for method users.}
		As the performance of a method mostly depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user’s existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method’s performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to their performance on a particular (set of) trajectory type. Further to the right are shown the accuracy (+: scaled performance $\geq$ 0.9, $\pm$: >0.6), usability scores (+:$\geq$0.9, $\pm$ $\geq$0.6), estimated running times and required prior information. k, thousands; m, millions.
	}
	\label{fig:figure_5}
\end{figure}

@@

\begin{figure}[tbh!]
	\centering\includegraphics[width=\linewidth]{fig/figure_6.pdf}
	\caption{
		\textbf{Demonstration of how a common framework for TI methods facilitates broad applicability using some example datasets.}
		Trajectories inferred by each method were projected to a common dimensionality reduction using multidimensional scaling. For each dataset, we also calculated a ‘consensus’ prediction, by calculating the cordist between each pair of models and picking the model with the highest score on average. \textbf{a}, The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC. \textbf{b}, The top methods applied on a dataset containing a bifurcating trajectory of reprogrammed fibroblasts. \textbf{c}, A synthetic dataset generated by dyntoy, containing four disconnected trajectories. \textbf{d}, A synthetic dataset generated by dyngen, containing a cyclic trajectory.
	}
	\label{fig:figure_6}
\end{figure}

@@

We gathered a list of 71 trajectory inference tools (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}{1}) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}. We welcome any contributions by creating an issue at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.

@@

We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}{2}). The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalized and filtered using a common pipeline, as discussed later. Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation. In the end, we included 110 datasets for this evaluation.

@@

The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{norel_selfassessmenttrapcan_2011}. Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer. We therefore defined several possible metrics for comparing trajectories, each investigating different layers. These are all discussed in Section~\ref{sec:dynb_supn1} along with examples and robustness analyses when appropriate.

Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (Section~\ref{sec:dynb_supn1}). Based on this analysis, we chose four metrics for the evaluation, each assessing a different aspect of the trajectory: (1) the HIM measures the topological similarity; (2) the F1branches compares the branch assignment; (3) the cordist assesses the similarity in pairwise cell–cell distances and thus the cellular positions; and (4) the wcorfeatures looks at whether similar important features (genes) are found in both the reference dataset and the prediction.

@@

The HIM metric \cite{jurman_himglocalmetric_2015} uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length). It is a linear combination of the normalized Hamming distance, which gives an indication of the differences in edge lengths, and the normalized Ipsen–Mikhailov distance, which assesses the similarity in degree distributions. The latter has a parameter $\gamma$, which was fixed at 0.1 to make the scores comparable between datasets. We illustrate the metric and discuss alternatives in Section~\ref{sec:dynb_supn1}.

@@

To compare branch assignment, we used an F1 score, also used used for comparing biclustering methods \cite{saelens_comprehensiveevaluationmodule_2018}. To calculate this metric, we first calculated the similarity of all pairs of branches between the two trajectories using the Jaccard similarity. Next, we defined the ‘Recovery' (respectively ‘Relevance') as the average maximal similarity of all branches in the reference dataset (respectively prediction). The F1branches was then defined as the harmonic mean between Recovery and Relevance. We illustrate this metric further in Section~\ref{sec:dynb_supn1}.

@@

To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics. This aggregation strategy is explained in more detail in Section~\ref{sec:dynb_supn1}.
